Static Analysis by Path Focusing

### 2011-05-06 report/report.tex
report

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{an abstract}
\keywords{keywords}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

\section*{Introduction}

@@

A fundamental fact in static analysis is that it cannot be perfectly precise:
either the set of possible states we compute is unsound (it doesn't contain all
the possible states during execution), or it is incomplete (it contains
states that can never be reached during execution). 
Abstract Interpretation is sound but incomplete, since it always computes an
upper-approximation of the set of possible states.

Linear Relation Analysis approximates the set of possible states as a convex
polyhedron where the dimensions are the numerical variables of the program.
Instead of convex polyhedra, one could choose intervals, octagons, etc.
The fixpoint equation is computed iteratively; that is, successive
approximations of the set of reachable states are computed until they converge
to a fixpoint.

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyzer, as
well as the implementation of another technique proposed in \cite{GopanR06},
such as we can compare precision and cost of the these two methods.
\end{itemize}

The report is organized as follows:
Part \ref{stateoftheartpart} introduces the state of the art in Abstract
Interpretation and more specifically in Linear Relation Analysis.
Part \ref{pathfocusingpart} describes the two techniques
\emph{Lookahead Widening} and \emph{Path Focusing}.
Part \ref{implementationpart} presents the implementation of these two
techniques into an analyzer, and the experimental results we obtained.

@@

Abstract Interpretation \cite{CC77,CousotCousot92-1} is a general method to find approximate solutions of
fixpoint equations. This method is used for program analysis, since analyzing a
program often comes down to solving fixpoint equations $\Phi(x) = x$.

@@

Abstract Interpretation method proposes to represent more efficiently the
elements of this complex domain $C$ of concrete values, by choosing an simpler
domain  $A$ called \emph{abstract domain}. The relation between these two
domains is characterized by two functions, usually called 
$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:

$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
\leq_{C} \gamma(y) $$

where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
fixpoint computation is done in the abstract domain, using the approximation of
the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
noted $\Phi^\#$.
We have the following result:

\begin{proposition}
If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
\end{proposition}

In fact, one compute the least fixpoint of the function in the abstract
domain, and the result of the computation also gives an upper approximation of
the fixpoint in the concrete domain.

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, or of finite depth, meaning that there doesn't exist
any sequence $(y_i)_{i \geq 0}$ of elements of the abstract domain $A$ such as
$\forall i \geq 0,\ y_i <_A y_{i+1}$. Indeed, in case of a domain of infinite
depth, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

@@

\begin{itemize}
\item $\forall y_1, y_2 \in A,\ y_1 \leq_A y_1 \nabla y_2$ and $y_2 \leq_A y_1
\nabla y_2$. This guarantees the correctness of the result.
\item For any increasing sequence $y_0 \leq_A y_1 \leq_A \dots$, the sequence
defined by 
$$\system{
y'_0 & = &  y_0 \\
y'_{i+1} & = &  y'_i\ \nabla\ y_{i+1},\ \  \forall i > 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

@@

The least fixpoint of a function $\Phi^\#$ is noted $\overline{y}$ and is equal
to $\displaystyle \lim_{i \geq 0}\ y_i$, where $y_0 = \perp$ (the least element
of $A$) and $y_{i+1} = \Phi^\# (y_i)$. Instead of computing this least fixpoint,
one compute an upper approximation of it, by computing the following ascending
sequence:

$$\system{
y'_0 &=& \perp \\
y'_{i+1} &=& y'_i\ \nabla\ \Phi^\#(y'_i),\ \ \forall i > 0
}$$

which converges towards $\tilde{y}$, where $\tilde{y} \geq_A \overline{y}$.

$\tilde{y}$ is a correct upper approximation of the least fixpoint of $\Phi^\#$,
and the classical technique is to regain some precision lost by the widening
operator by computing a descending sequence:

$$\system{
y''_0 &=& \tilde{y} \\
y''_{i+1} &=& \Phi^\#(y''_i), \ \ \forall i > 0
}$$

Each element of this descending sequence is still an upper approximation of the
least fixpoint $\overline{y}$. So, this sequence often allows to find
approximate least fixpoints that are more precise than the one obtained after
the ascending sequence.

@@

To each state $p \in P$ of the control flow graph, we associate an abstract
value $X_p \in D$, $D$ being in our case the domain of convex polyhedra over
$\Q^n$. Since the exact operation $\tau$ may not be expressed in the abstract
domain, we abstract it into $\tau^\#$ such as $\forall X \in D, \tau(X)
\subseteq \tau^\#(X)$.

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such as removing the nodes of $P_w$
disconnects every cycles. For instance, one could choose $P_w$ as the set of the
heads of loops.

@@

Lookahead widening \cite{GopanR06} is a technique that aims to isolate the
different loop phases during the analysis, in order to obtain more precise
results.

@@

@@

	Let $(P,E)$ be the control flow graph of the program.
	First, one choose a set of points $P_W \subseteq P$, such that the graph
	obtained after removing these points has no cycle. The widening operator
	will be applied at these points. The choice of $P_W$ can be guided by
	\cite{Bou92}.
	Then, one choose another set $P_R \subseteq P$ of nodes, satisfying the
	following properties:
	\begin{itemize}
	\item $I_p = \emptyset$ for each $p \in P \setminus P_R$, meaning that the initial
	points of the program are included in $P_R$.
	\item $P_W \subseteq P_R$: all the widening points are in $P_R$.
	\end{itemize}

	The multigraph is then the graph such as:
	\begin{itemize}
	\item each element of $P_R$ is a point of the graph.
	\item for each path $p_1 \rightarrow p_2 \rightarrow \dots \rightarrow p_k$
	in the control flow graph, such as $p_1 \in P_R, p_k \in P_R$,
	there is a transition in the multigraph from $p_1$ to $p_k$ with the
	semantic $$\tau_{p_1 \rightarrow \dots \rightarrow p_k} = \tau_{p_{k-1}
	\rightarrow p_k} \circ
	\dots \circ \tau_{p_1 \rightarrow p_2}$$ where $\tau_{p_i \rightarrow
	p_{i+1}}$ is the transformation associated to the semantic of $p_i
	\rightarrow p_{i+1}$.
	\end{itemize}

@@

  \path [transition] 
		(n0) edge              node [yshift=5mm] {$x,y \leftarrow 0$} (n1);
  \path [transition] 
		(n1) edge    [bend left]  node [right,yshift=-0.8cm]  {$
		\begin{array}{l}
			x \geq 51 \\
			y \leq 0 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
  \path [transition2] 
		(n1) edge    [loop left,distance=1.2cm]  node [left,xshift=3mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \geq -1 \\
			y \leftarrow y+1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [loop right,distance=1.2cm]  node [right,xshift=-2mm] {$
		\begin{array}{l}
			x \geq 51 \\
			y \geq 1 \\
			y \leftarrow y-1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [bend right]  node [left,yshift=-0.8cm,xshift=4mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \leq -2 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
\end{tikzpicture}
\end{minipage}
\caption{Example of classical control flow graph, and its associate expanded
multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
in the control flow graph corresponds to a simple dashed transition in the multigraph,
having the same semantic.}
\end{figure}
\FloatBarrier

@@

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the Boolean variables such that the formula is true.
	Among these Boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the Boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path doesn't go through this transition.
	\end{itemize}

@@

(b) control flow graph (a), such as it is represented by the SMT formula $\rho$.
Points in $P_R$ have been split into a ``source'' point and a ``destination''
point, and $\Phi$-variables associated to points in $P_R$ have been primed.
\end{minipage} 
\caption{Example (cont'd)}
\end{figure}

@@

	\begin{itemize}
	\item we search a path starting in $p_i$: $b_i^s$ has to be \emph{true}, and
	all other source points are \emph{false}.
	\item the vector $x_i$ of the numerical variables at point $p_i$ is in our
	abstract value $X_{p_i}$
	\item we search a path arriving in a successor $p_j$ of $p_i$, such as the
	vector of the numerical variables $x_j'$ is not yet included in the current
	abstract value $X_{p_j}$ of $p_j$. $b_j^d$ is \emph{true}, and the
	conjunction of the inequalities given by $X_{p_j}$ is \emph{false}.
	\end{itemize}

@@

@@

@@

	\begin{definition}[Liveness by linearity]
	A variable $v$ is \emph{live by linearity} at the control point $p$ iff:
	\begin{itemize}
	\item 
	its value is available at $p$, meaning that there is a path from the
	point defining the variable to $p$,
	\item one of these conditions holds:
		\begin{itemize}
		\item $v$ is live in $p$.
		\item There is a variable $v'$, defined as a linear combination of other
		variables $v_1, v_2, \dots, v_n$, such that $\exists i \in \{1,\dots,n\}, v = v_i$,
		and $v'$ is live by linearity in $p$.
		\end{itemize}
	\end{itemize}
	\end{definition}

@@

	\subsection{SMT-solving}

	\emph{Path focusing} technique requires to decide the satisfiability of some
	SMT formula, and to extract a model when the formula is satisfiable, i.e an
	assignement of the variables such that the formula is \emph{true}. For doing
	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
	and Yices \cite{DutertreM06}. One can easily switch from one to another with
	an argument we give as parameter to the analyzer. In this way, we can
	compare the effectiveness of both SMT-solvers.

@@

  \section*{Conclusion}

@@

\bibliographystyle{plain}
  \bibliography{report}
\end{document}

>>>>

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification aims at statically discover properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
	\\ \\  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation, Invariant
Generation}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

\part{Introduction}

@@

A fundamental fact in static analysis is that it cannot be perfectly precise:
either the set of possible states we compute is unsound (it doesn't contain all
the possible states during execution), or it is incomplete (it contains
states that can never be reached during execution). 
Abstract Interpretation is sound but incomplete, since it always computes an
upper-approximation of the set of possible states. The challenge is then to be
as precise as possible, so that we can prove properties about the program.

Linear Relation Analysis over-approximates the set of possible states as a
convex polyhedron where the dimensions are the numerical variables of the
program.
Instead of convex polyhedra, one could choose intervals, octagons, etc.
The fixpoint equation is computed iteratively; that is, successive
approximations of the set of reachable states are computed until they converge
to a fixpoint.

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyzer, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of the these two methods.
\end{itemize}

The report is organized as follows:
Part \ref{stateoftheartpart} introduces the state of the art in Abstract
Interpretation and more specifically in Linear Relation Analysis.
Part \ref{pathfocusingpart} describes the two techniques
\emph{Lookahead Widening} and \emph{Path Focusing}.
Part \ref{implementationpart} presents the implementation of these two
techniques into an analyzer, and the experimental results we obtained.
Finally, Part \ref{future} introduces some directions to explore.

@@

Abstract Interpretation \cite{CC77,CousotCousot92-1} is a general method for
finding approximate solutions of
fixpoint equations. This method is used for program analysis, since analyzing a
program often comes down to solving fixpoint equations, because of the loops and
recursive procedures.

@@

Abstract Interpretation method proposes to represent more efficiently the
elements of this complex domain $C$ of concrete values,
by choosing an simpler domain  $A$ called \emph{abstract domain}.

A concretization function $\gamma$ maps the elements of $A$ to elements of $C$:
\begin{eqnarray*}
\gamma: & A &\longrightarrow C \\
		& x &\longmapsto \gamma(x)
\end{eqnarray*}

The concrete semantics is a monotonic function $\Phi: C \rightarrow C$.
One can choose an abstract semantics $\Phi^\#: A \rightarrow A$ satisfying this
condition:

$$\forall x \in A, \Phi \circ \gamma (x) \subseteq_C \gamma \circ \Phi^\#(x)$$

$\Phi^\#$ is then an abstraction of $\Phi$, that gives an upper
approximation of $\gamma(x)$ when composed with $\gamma$.

\begin{proposition}
Suppose $C$ is a complete lattice, and $\Phi$ is increasing from $C$ to $C$.
Each $x \in A$, satisfying the condition $\Phi^\#(x) = x$, is an abstraction of
the least fixpoint of $\Phi$.
\end{proposition}

Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique aims at computing
one of these $x$. It computes the stationary limit of an ascending sequence
$(x_i)_{i \geq 0}$ defined by the induction:

$$\left[
\begin{array}{lll}
x_0 &=& \perp \\
x_{n+1} &=& \Phi^\#(x_n)
\end{array}
\right.
$$
%The relation between these two
%domains is characterized by two functions, usually called 
%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
%
%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
%\leq_{C} \gamma(y) $$
%
%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
%fixpoint computation is done in the abstract domain, using the approximation of
%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
%noted $\Phi^\#$.
%We have the following result:
%
%\begin{proposition}
%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
%\end{proposition}
%
%In fact, one compute the least fixpoint of the function in the abstract
%domain, and the result of the computation also gives an upper approximation of
%the fixpoint in the concrete domain.

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, or of finite depth, meaning that there doesn't exist
any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, in case of a domain of infinite
depth, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

@@

\begin{itemize}
\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \nabla x_2$ and $x_2 \leq_A x_1
\nabla x_2$. This guarantees the correctness of the result.
\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
defined by 
$$\system{
x'_0 & = &  x_0 \\
x'_{i+1} & = &  x'_i\ \nabla\ x_{i+1},\ \  \forall i > 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

@@

The least fixpoint of a function $\Phi^\#$ is noted $\overline{x}$ and is equal
to $\displaystyle \lim_{i \geq 0}\ x_i$, where $x_0 = \perp$ (the least element
of $A$) and $x_{i+1} = \Phi^\# (x_i)$. Instead of computing this least fixpoint,
one compute an upper approximation of it, by computing the following ascending
sequence:

$$\system{
x'_0 &=& \perp \\
x'_{i+1} &=& x'_i\ \nabla\ \Phi^\#(x'_i),\ \ \forall i > 0
}$$

which converges towards $\tilde{x}$, where $\tilde{x} \geq_A \overline{x}$.

$\tilde{x}$ is a correct upper approximation of the least fixpoint of $\Phi^\#$,
and the classical technique is to regain some precision lost by the widening
operator by computing a descending sequence:

$$\system{
x''_0 &=& \tilde{x} \\
x''_{i+1} &=& \Phi^\#(x''_i), \ \ \forall i > 0
}$$

Each element of this descending sequence is still an upper approximation of the
least fixpoint $\overline{x}$. So, this sequence often allows to find
approximate least fixpoints that are more precise than the one obtained after
the ascending sequence.

@@

To each state $p \in P$ of the control flow graph, we associate an abstract
value $X_p \in D$, $D$ being in our case the domain of convex polyhedra over
$\Q^n$. Since the exact operation $\tau$ may not be expressed in the abstract
domain, we abstract it into $\tau^\#$ such that $\forall X \in D, \tau(X)
\subseteq \tau^\#(X)$.

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
disconnects every cycles. For instance, one could choose $P_w$ as the set of the
heads of loops.

@@

Lookahead widening \cite{GopanR06} is a technique that isolates the
different loop phases during the analysis, and stabilizes each of these loop
phases before proceeding to the next. One only consider paths that are feasible
at the first iteration, and forget the others for a while. The iterations end
when the analysis of these paths has converged. After that, one can compute some
narrowing iterations. Then, paths that
become feasible are reinserted into the control flow graph, and we redo
iterations.

@@

In the classical abstract interpretation technique, narrowing iterations are
computed at the end of the analysis, when all the fixpoint computation has
converged.
The interest of this method is to apply narrowing iterations right after each
path computation. This allows to recover precision before analyzing new paths,
and then to be more precise.

@@

	Let $(P,E)$ be the control flow graph of the program.
	First, one choose a set of points $P_W \subseteq P$, so that the graph
	obtained after removing these points has no cycle. The widening operator
	will be applied at these points. The choice of $P_W$ can be guided by
	\cite{Bou92}.
	Then, one choose another set $P_R \subseteq P$ of nodes, satisfying the
	following properties:
	\begin{itemize}
	\item $I_p = \emptyset$ for each $p \in P \setminus P_R$, meaning that the initial
	points of the program are included in $P_R$.
	\item $P_W \subseteq P_R$: all the widening points are in $P_R$.
	\end{itemize}

	The multigraph is then the graph such that:
	\begin{itemize}
	\item each element of $P_R$ is a point of the graph.
	\item for each path $p_1 \rightarrow p_2 \rightarrow \dots \rightarrow p_k$
	in the control flow graph, such that $p_1 \in P_R, p_k \in P_R$,
	there is a transition in the multigraph from $p_1$ to $p_k$ with the
	semantic $$\tau_{p_1 \rightarrow \dots \rightarrow p_k} = \tau_{p_{k-1}
	\rightarrow p_k} \circ
	\dots \circ \tau_{p_1 \rightarrow p_2}$$ where $\tau_{p_i \rightarrow
	p_{i+1}}$ is the transformation associated to the semantic of $p_i
	\rightarrow p_{i+1}$.
	\end{itemize}

@@

  \path [transition] 
		(n0) edge              node [yshift=5mm] {$x,y \leftarrow 0$} (n1);
  \path [transition] 
		(n1) edge    [bend left]  node [right,yshift=-0.8cm]  {$
		\begin{array}{l}
			x \geq 51 \\
			y \leq 0 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
  \path [transition2] 
		(n1) edge    [loop left,distance=1.2cm]  node [left,xshift=3mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \geq -1 \\
			y \leftarrow y+1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [loop right,distance=1.2cm]  node [right,xshift=-2mm] {$
		\begin{array}{l}
			x \geq 51 \\
			y \geq 1 \\
			y \leftarrow y-1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [bend right]  node [left,yshift=-0.8cm,xshift=4mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \leq -2 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
\end{tikzpicture}
\end{minipage}
\caption{Example of classical control flow graph, and its associate expanded
multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
in the control flow graph corresponds to a simple dashed transition in the multigraph,
having the same semantic.}
\label{multigraphtransformGopan}
\end{figure}
\FloatBarrier

@@

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the Boolean variables so that the formula is true.
	Among these Boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the Boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path doesn't go through this transition.
	\end{itemize}

@@

(b) control flow graph (a), such as it is represented by the SMT formula $\rho$.
Points in $P_R$ have been split into a ``source'' point and a ``destination''
point, and $\Phi$-variables associated to points in $P_R$ have been primed.
\end{minipage} 
\caption{Example (cont'd)}
\label{multigraphGopan}
\end{figure}

@@

	\begin{itemize}
	\item we search a path starting in $p_i$: $b_i^s$ has to be \emph{true}, and
	all other source points are \emph{false}.
	\item the vector $x_i$ of the numerical variables at point $p_i$ is in our
	abstract value $X_{p_i}$
	\item we search a path arriving in a successor $p_j$ of $p_i$, such that the
	vector of the numerical variables $x_j'$ is not yet included in the current
	abstract value $X_{p_j}$ of $p_j$. $b_j^d$ is \emph{true}, and the
	conjunction of the inequalities given by $X_{p_j}$ is \emph{false}.
	\end{itemize}

@@

	When the self loop is accelerable \cite{Gon07}, instead of doing a
	widening/narrowing sequence, one can directly accelerate the loop and find
	its transitive closure.

@@

	\subsection{Example}

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. One computes
	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.

\begin{itemize}
\item Step 0: Initially, $X_0 = \top, X_1 = \perp, X_7 = \perp$, and $A =
\{p_0\}$.
\item Step 1: We start in $p_0$. Is there a path starting in $p_0$, that arrives
in a state in $P_R$ and makes its abstract value grow? The SMT-solver answers
\emph{yes}, and gives the path $p_0 \rightarrow p_1$. We update $X_1$:
it is the polyhedron $x=y=0$. We update $A = \{p_1\}$. 
We do another SMT query: is there another path
starting in $p_0$ that makes an abstract value grow? Answer is no.
\item Step 2: We now work on $p_1$. Is there a path starting in $p_1$ that
arrives in a successor of $p_1$, and makes its abstract value grow? Answer is
yes, and the model given by the SMT-solver is the path $p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. Actually, this
is the first phase of the loop. 
The image of $x=y=0$ by this transition is $x=y=1$. The new $X_1$ is then $x=y
\wedge 0\leq x \leq 1$. The path is a self loop, but is seen for the first time,
so we do not apply widening yet. So, we redo an SMT-query, that gives us exactly
the same path. Then, we can do a widening/narrowing sequence. 
After widening, we have $x=y \wedge x \geq 0$, and narrowing gives $X_1 = (x=y)
\wedge (0 \leq x \leq 51)$.
\item Step 3: We still are in $p_1$: we ask for another focus path, and the
SMT-solver finds the path $p_1 \rightarrow p_3
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. This is the
second phase of the loop, which is possible since $(x,y)=(51,51) \in X_1$.
Again, this is a self-loop, but we do not widen yet. The
image of $X_1$ by the transition is the single point $(52,50)$. $X_1$ is then
the convex hull of this point with the old $X_1$. A new SMT query gives the same
path, and we can now apply widening/narrowing. We obtain
$X_1 = (x \leq y) \wedge (102 - x - y \geq 0)$ after widening, and narrowing
gives $X_1 = (x \leq y) \wedge (102 - x - y \geq 0) \wedge y \geq 0$.
We can see that $X_1$ is the smallest polyhedron containing the possible values
for $(x,y)$ (see Figure \ref{result}.b).
\item Step 4: Again, we ask the SMT-solver for a new focus path, and it returns
$p_1 \rightarrow p_3 \rightarrow p_4 \rightarrow p_7$. The image of the
polyhedra $X_1$ after this path transformation is the single point $y=-1 \wedge
x = 102$. We can update $X_7 = {(102,-1)}$. Path focusing technique gives us the
precise result of this program. After that, the SMT-solver can't find any new
path, so the analysis terminates.
\end{itemize}

@@

	\begin{definition}[Liveness by linearity]
	A variable $v$ is \emph{live by linearity} at the control point $p$ iff:
	\begin{itemize}
	\item 
	its value is available at $p$, meaning that there is a path from the
	point defining the variable to $p$,
	\item one of these conditions holds:
		\begin{itemize}
		\item $v$ is live in $p$.
		\item There is a variable $v'$, defined as a linear combination of other
		variables $v_1, v_2, \dots, v_n$, so that $\exists i \in \{1,\dots,n\}, v = v_i$,
		and $v'$ is live by linearity in $p$.
		\end{itemize}
	\end{itemize}
	\end{definition}

@@

	\section{SMT-solving}

	\emph{Path focusing} technique requires to decide the satisfiability of some
	SMT formula, and to extract a model when the formula is satisfiable, i.e an
	assignement of the variables so that the formula is \emph{true}. For doing
	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
	and Yices \cite{DutertreM06}. One can easily switch from one to another with
	an argument we give as parameter to the analyzer. In this way, we can
	compare the effectiveness of both SMT-solvers.

@@

	To be completed\dots

\part{Future Work}
\label{future}
\emph{Path Focusing} technique allows to work on control flow graphs with a huge
number of paths, since they are computed only when needed. With this in mind,
one could think of several improvements using path focusing, by encoding into
the control flow graph new informations as paths, such as arithmetic overflows,
alias analysis, \dots
In this part, we explain how to encode such kind of informations into implicit
paths of the CFG.

	\section{Arithmetic Overflows}

Some operations such as additions, soustractions\dots may overflow if their
result is not in the type's definition interval. For instance, an
\emph{int} variable can only take values in $[-2^{31}, 2^{31} -1]$. Then, there
are two kind of overflows. Suppose $N=2^{31}$:
\begin{itemize}
\item The result $r$ of the operation may be greater than $N-1$. In this case,
the result we obtain is actually $r-2N$.
\item $r$ may be lesser than $-N$. Then, the result is $r+2N$.
\end{itemize}

The idea would be to distinguish the correct behaviour and the two kind of
overflow in the control flow graph, each behaviour corresponding to a specific
path (see Figure \ref{overflow}).

\begin{figure}[!h]
\centering
\begin{tikzpicture}[->,>=stealth',auto,node distance=3cm,
                    semithick,font=\footnotesize]
	\node[state] (n0) {};
	\node[state] (n1) [below of=n0] {};

	\path [transition] 
		(n0) edge [out=-140,in=140]  node [xshift=-0.2mm] 
		{$\begin{array}{l} x+y < -N\ /\\ r=x+y+2N\end{array}$} (n1)
		(n0) edge [in=170,out=190,distance=2cm]  node [left]
		{$\begin{array}{l} x+y > N-1\ / \\ r=x+y-2N\end{array}$} (n1)
		(n0) edge [in=10,out=-10,distance=2.5cm]  node [right] 
		{$\begin{array}{l} -N \leq x+y \leq N-1\ /\\ r=x+y\end{array}$} (n1)
		;
\end{tikzpicture}
\caption{How to treat overflows of the addition $x+y$}
\label{overflow}
\end{figure}

This results in a huge blowup in the number of paths, but hopefully a lot of
them will never be possible, so they will never be computed by our algorithm.

	\section{Alias analysis}

One could use the same idea as for overflows, in order to treat aliases.
Assume a pointer $p$ that possibly points to variables $x,y,z$ or $t$. For each
operation involving $*p$, one distinguishes these different cases by
creating a specific path for each of them. 
If the pointers are manipulated in a tricky way, one considers they may
point to each variable of the program. Again, there is an exponential blowup in
the number of paths, but these are treated implicitly.

\begin{figure}[!h]
\centering
\begin{tikzpicture}[->,>=stealth',auto,node distance=3cm,
                    semithick,font=\footnotesize]
	\node[state] (n0) {};
	\node[state] (n1) [below of=n0] {};

	\path [transition] 
		(n0) edge [in=170,out=190,distance=2.5cm]  node [left] 
		{$\begin{array}{l}p=\&x/ \\ x++\end{array}$} (n1);
	\path [transition] 
		(n0) edge  [in=120, out=-120] node [left,xshift=0.2cm] 
		{$\begin{array}{l}p=\&y/ \\ y++\end{array}$} (n1);
	\path [transition] 
		(n0) edge [in=60,out=-60]  node  [right,xshift=-0.2cm]
		{$\begin{array}{l}p=\&z/ \\ z++\end{array}$} (n1);
	\path [transition] 
		(n0) edge [in=10,out=-10,distance=2.5cm]  node [right] 
		{$\begin{array}{l}p=\&t/ \\ t++\end{array}$} (n1);
\end{tikzpicture}
\caption{How to treat pointers: $(*p)++$}
\label{pointers}
\end{figure}

\part{Conclusion}

@@

\bibliographystyle{alpha}
  \bibliography{report}
\end{document}
